Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-14 doc/invt.tex
invt.tex: update to Sec. II.C

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing
  \note{``incrementally building up''?}%
  a bias potential to offset
  the potential of mean force (PMF).
  %
  The two differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bias potential of
  the bin containing the current $z$,
  while metadynamics updates also
  those of neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the final bias potential,
  hence that of the PMF,
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general multiple-bin updating scheme.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  The optimal schedule of these schemes
  is given by the inverse-time formula,
  with a shifted origin of time such that
  the initial updating magnitude is half of
  the previous equilibrium value.
\end{abstract}

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which % brings about a side effect of disrupting
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

@@

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case,
and gives the optimal value of the origin of time.
%
It also predicts, for metadynamics,
a more complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, we generalize
the single-bin updating scheme
used by the WL algorithm
to a class of multiple-bin updating schemes
that are optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

For later convenience,
we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
the updates to $v_i$'s are
the same as the updates to $V_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should ideally be a constant of $i$
such that $\pi_i \approx p_i$.

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where,
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
  \label{eq:error_asym1}
\end{align}
%
\note{The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\Err$
that is about $\alpha_{\max}^2$ times as large,
where $\alpha_{\max} = \max \alpha(t)$.
%
This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_k(\omega) = \Gamma_k
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_k(\omega) = \Gamma_k + \Gamma^{(2)}_k \omega^2 + \dots,
  $$
  which leads to a correction to $\Err$
  $$
  \Gamma^{(2)}_k
  \int_0^T \ddot u_k^2\bigl( q(t) \bigr) \, dt
  \propto
  \alpha^2 \, \Err.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_asym1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
}
%
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{k = 1}^n
      \Gamma_k \, \lambda_k \,
      u_k\bigl( q(t) \bigr) \,
      \ddot u_k\bigl( q(t) \bigr) = 0
    .
    \tag{N2}
    \label{neq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_k$ as the position,
  and $\Gamma_k \, \lambda_k$ as the mass,
  then the right-hand side gives the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
In this way,
the underlying sampling process
affects the error only through the set of numbers, $\Gamma_i$'s.

@@

We evaluate the integrals of the autocorrelation functions,
$\Gamma_k$'s,
in two special cases.
%
First, if the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_k = 1.
  \label{eq:Gamma_perfect}
\end{equation}
%
In this case,
the bin index $i$ is an independent random variable
at each time step $t$, and
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \, \delta(t - t'),
  \label{eq:zz_perfect}
\end{equation}
%
where $\delta(t - t')$ is the equivalent of $\delta_{t, t'}$
in the continuous limit.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \left\langle
    \zeta_i(t)
  \right\rangle
  \left\langle
    \zeta_j(t')
  \right\rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus,
for $k \ge 2$, we get
%
\begin{equation}
  \kappa_k(t - t')
  =
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \langle \zeta_j(t) \, \zeta_l(t') \rangle
  %
  = \delta(t - t'),
\notag
%\label{eq:kappa_perfect}
\end{equation}
%
where we have used
Eqs. \eqref{eq:eig_orthonormal_rows},
\eqref{eq:eigenmode1},
\eqref{eq:eta_def},
\eqref{eq:zz_perfect},
and
$$
  \sum_{j = 1}^n \phi_{kj}
  =
  \sum_{j = 1}^n \frac{ \phi_{kj} \, \phi_{1j} } { p_j }
  =
  \delta_{k1}
  =
  0.
$$
%
\note{Proof.%Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_k(t - t')
  &=
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \langle \zeta_j(t) \, \zeta_l(t') \rangle
  \\
  %
  &=
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \left(
    p_j \, \delta_{j, l} - p_j \, p_l
  \right)
  \, \delta(t - t')
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{kj}^2 } { p_j }
    -
    \sum_{j = 1}^n \phi_{kj}
    \sum_{l = 1}^n \phi_{kl}
  \right)
  \, \delta(t - t')
  \\
  &=
  (1 - 0 \cdot 0) \, \delta(t - t').
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
}

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf Q
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
      \frac {1 - \sigma} 2 & \dots & 0 & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For a large class of updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf Q$ and $\mathbf w$
share eigvenvectors, but not eigenvalues.
%
The eigenvalues of $\mathbf Q$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value
$\kappa_k(0) = 1$
being the same as
that in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
This case is intended to model a local sampling process.

\subsection{\label{sec:mbin_opta}
Optimal schedule for the multiple-bin updating scheme}

@@

Before addressing the optimal $\alpha(t)$,
we shall first consider a constant schedule
$\alpha(t) \equiv \alpha_0$,
and compute the equilibrium error, $\Err$,
in this case.
%
After a long period $T'$,
the residual error becomes negligible, and
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^{T'}
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &
  \stackrel{ T' \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0
  .
\notag
%\label{eq:error_eql}
\end{align}
%
In terms of the components, we have
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:y2_eql}
\end{equation}
%

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, for $t < 0$, we assume
$\alpha(t) = \alpha_0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k(t) = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on the schedule $\alpha(t)$
only through endpoint value of
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate the schedule under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

The idea is to measure $\Gamma_k$'s
via a simulation with a constant schedule,
$\alpha(t) = \alpha_0$,
using Eq. \eqref{eq:y2_eql}.
%
A practical difficulty is that
we often do not know the intrinsic distribution, $p^*_i$,
beforehand, and thus cannot compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we shall use
the following modification:
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\\
\bar V
&=
\sum_{ j = 1 }^n p_j \, V_j.
\notag
\end{align}

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing
  \note{``incrementally building up''?}%
  a bias potential to offset
  the potential of mean force (PMF).
  %
  The two differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bias potential of
  the bin containing the current $z$,
  while metadynamics updates also
  those of neighboring bins according to
  a Gaussian window.
  %
  The error of the final bias potential,
  hence that of the PMF,
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general multiple-bin updating scheme.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  The optimal schedule of these schemes
  is given by the inverse-time formula,
  with a shifted origin of time such that
  the initial updating magnitude is half of
  the previous equilibrium value.
\end{abstract}

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian window
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which % brings about a side effect of disrupting
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

@@

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case,
and gives the optimal value of the origin of time.
%
It also predicts, for metadynamics,
a more complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the single-bin updating scheme
used by the WL algorithm
to a class of multiple-bin updating schemes
that are optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

For later convenience,
we introduce a shifted bias potential
%
\begin{equation}
  v_i
  \equiv
  V_i
  +
  \ln \frac { p_i }
            { p^*_i }
  .
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
the updates to $v_i$'s are
the same as the updates to $V_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should ideally be a constant of $i$
such that $\pi_i \approx p_i$.

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
  \label{eq:error_asym1}
\end{align}
%
\note{The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\Err$
that is about $\alpha_{\max}^2$ times as large,
where $\alpha_{\max} = \max \alpha(t)$.
%
This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_k(\omega) = \Gamma_k
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_k(\omega) = \Gamma_k + \Gamma^{(2)}_k \omega^2 + \dots,
  $$
  which leads to a correction to $\Err$
  $$
  \Gamma^{(2)}_k
  \int_0^T \ddot u_k^2\bigl( q(t) \bigr) \, dt
  \propto
  \alpha^2 \, \Err.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_asym1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
}%
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{k = 1}^n
      \Gamma_k \, \lambda_k \,
      u_k\bigl( q(t) \bigr) \,
      \ddot u_k\bigl( q(t) \bigr) = 0
    .
    \tag{N2}
    \label{neq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_k$ as the position,
  and $\Gamma_k \, \lambda_k$ as the mass,
  then the right-hand side gives the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}%
In this way,
the underlying sampling process
affects the error only through the set of numbers, $\Gamma_i$'s.

@@

We can evaluate the integrals of the autocorrelation functions,
$\Gamma_k$'s,
in two special cases.
%
First, if the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_k = 1.
  \label{eq:Gamma_perfect}
\end{equation}
%
To see this, we note that
the bin index $i$ is an independent random variable
at each time step $t$ in this case, and
%
\begin{align}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle
  \langle h_j(t') \rangle
  \notag
  \\
  &=
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \,
  \delta_{t, t'}
  ,
  \label{eq:zz_perfect}
\end{align}
%
where $\delta(t - t')$ is the equivalent of $\delta_{t, t'}$
in the continuous limit.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \left\langle
    \zeta_i(t)
  \right\rangle
  \left\langle
    \zeta_j(t')
  \right\rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus,
for $k \ge 2$, we get
%
\begin{equation}
  \kappa_k(t - t')
  =
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  %
  = \delta_{t, t'},
\notag
%\label{eq:kappa_perfect}
\end{equation}
%
where we have used
Eqs. \eqref{eq:eig_orthonormal_rows}-\eqref{eq:ortho1},
\eqref{eq:eta_def},
and
\eqref{eq:zz_perfect}.
%
\note{Proof.%Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_k(t - t')
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  \\
  %
  &=
  \sum_{i,j = 1}^n
  \frac{ \phi_{ki} } { p_i }
  \frac{ \phi_{kj} } { p_j }
  \left(
    p_i \, \delta_{i, j} - p_i \, p_j
  \right)
  \, \delta_{t, t'}
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{kj}^2 } { p_j }
    -
    \sum_{i = 1}^n \phi_{ki}
    \sum_{j = 1}^n \phi_{kj}
  \right)
  \, \delta_{t, t'}
  \\
  &=
  (1 - 0 \cdot 0) \, \delta_{t, t'}.
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
}

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf Q
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf Q$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf Q$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}
%

\subsection{\label{sec:mbin_opta}
Optimal schedule for multiple-bin updating schemes}

@@

Before addressing the optimal schedule
for a multiple-bin scheme,
we shall first compute the equilibrium error
under a constant schedule,
$\alpha(t) = \alpha_0$.
%
After a long period $T'$,
the residual error becomes negligible, and
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^{T'}
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &
  =
  %\stackrel{ T' \to \infty }
  %{ =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0
  ,
  \qquad
  ( T' \to \infty )
  .
\notag
%\label{eq:error_eql}
\end{align}
%
In terms of the components, we have
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:y2_eql}
\end{equation}
%

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, we assume
$\alpha(t) = \alpha_0$
for $t < 0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k(t) = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on $\alpha(t)$ only through
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate $\alpha(t)$ under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

The idea is to run
a simulation with a constant schedule,
$\alpha(t) = \alpha_0$,
and then use Eq. \eqref{eq:y2_eql}.
%
A practical difficulty is that
we often do not know the intrinsic distribution, $p^*_i$,
in advance to compute
the shifted bias potential, $v_i$,
from the original one, $V_i$
(cf. Sec. \ref{sec:FDS}).
%
To circumvent the problem, we shall
modify Eq. \eqref{eq:y2_eql} as
%
\begin{equation}
  \operatorname{var} Y_k
  =
  \frac{1}{2}
  \alpha \, \Gamma_k \, \lambda_k,
\label{eq:varY}
\end{equation}
%
where
``$\operatorname{var}$''
means the variance from time averaging,
and
%
\begin{align}
Y_k
&=
\sum_{ i = 1 }^n
  \phi_{k i} \, (V_i - \bar V),
\label{eq:Y_def}
\\
\bar V
&=
\sum_{ j = 1 }^n p_j \, V_j.
\notag
\end{align}

To show Eq. \eqref{eq:varY},
we first use Eq. \eqref{eq:v_def}
to rewrite Eq. \eqref{eq:x_def}
as
$$
\begin{aligned}
  x_i
  &=
  (V_i - \bar V) + s_i,
  \\
  s_i
  &=
  \ln \frac{ p_i } { p^*_i }
  -
  \sum_{ j = 1 }^n
    p_j \, \ln \frac{ p_j } { p^*_j }.
\end{aligned}
$$
Thus, by Eqs. \eqref{eq:y_def} and \eqref{eq:Y_def},
we have
$$
y_k = Y_k + \sum_{i = 1}^n \phi_{k i} \, s_i.
$$
Since the second term on the right-hand side
is a constant, and the average of $y_k$ is zero,
we have
$$
\operatorname{var} Y_k
=
\operatorname{var} y_k
=
\langle y_k^2 \rangle.
$$
Finally, by using Eq. \eqref{eq:y2_eql},
we reach Eq. \eqref{eq:varY}.
